---
title: "Work Task Seventeen: Implement a DataSHIELD compatible zCompositions::lrEM"
authors: "Xavier Escribà Montagut & Stuart Wheater"
format:
  html:
    theme: cosmo   # You can also try: flatly, lumen, journal, etc.
    toc: true
    code-overflow: wrap
    smooth-scroll: true
    page-layout: full
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
  cache.refresh: false
  freeze: auto
---

## Discussion of the work task

The work task involves implementing a DataSHIELD compatible version of the zCompositions::lrEM function from the zCompositions R package.


## What does zCompositions::lrEM do?

The `zCompositions::lrEM` function implements a log-ratio Expectation-Maximization algorithm for imputing missing values, zeros, and values below detection limits in compositional data. It uses log-ratio transformations (ALR/ILR) to work in Euclidean space while preserving compositional properties. The function is particularly useful for environmental, geochemical, and microbiome data where compositional datasets frequently contain structural zeros or censored observations below detection thresholds. (Documentation: https://cran.r-project.org/web/packages/zCompositions/zCompositions.pdf)

## How is zCompositions::lrEM implemented as a DataSHIELD compatible function?

The zCompositions::lrEM function is implemented as a DataSHIELD assign function (lrEMDS) since it needs to operate on data that resides on the analysis server. It takes a table as input, processes it on the server-side to create a new data.frame with the imputed values, and returns that object of data.frame class.

## Solution

A 2-stage assign DataSHIELD function has been developed and placed into dsCoda. This function acts as a wrapper to the target function zCompositions::lrEM.

## Test script

### Library Setup

We begin by loading all the required packages for the DataSHIELD analysis. This includes the core DataSHIELD infrastructure packages (`DSI`, `DSOpal`), client packages for base functions (`dsBaseClient`), tidyverse operations (`dsTidyverseClient`), compositional analysis (`dsCodaClient`), and survival analysis (`dsSurvivalClient`).

```{r setup, cache=FALSE}
# load the required packages
library('DSI')  
library('DSOpal')
library('dsBaseClient') # Make sure this is v6.3.2-dev `devtools::install_github("datashield/dsBaseClient", ref = "v6.3.2-dev")`
library('dsTidyverseClient')
library('dsSurvivalClient') # Make sure this is v6.3.2-dev `devtools::install_github("datashield/dsSurvivalClient", ref = "v2.3.0-dev")`
library('dsCodaClient') # Make sure this is task-17 branch `devtools::install_github("datashield/dsCodaClient", ref = "task-17")`
```

### Data Preparation

#### Local Data Processing

We start by reading the synthetic lifestyle data and preparing it for upload to the Opal server. The synthetic data is stored in the `lifestyle_study1.rds`, `Sim_df_accel.rds`, and `Sim_df_mort.rds` files. A new data.frame is created by merging the three files on the `ID` variable.

```{r data-prep-local, cache=FALSE}
lifestyle_study1 <- readRDS("data/lifestyle_study1.rds")
df_actipass <- readRDS("data/Sim_df_accel.rds")
df_mort <- readRDS("data/Sim_df_mort.rds")
df <- merge(df_actipass, lifestyle_study1, by=c("ID"))
df <- merge(df, df_mort, by=c("ID"))

save(df, file='df.rda')
```

#### Opal Server Setup

We define a helper function for uploading datasets to the Opal server that handles both CSV and RDA file formats. The function creates projects if they don't exist and properly formats data with row identifiers for DataSHIELD compatibility.

After defining the function, we log into the Opal demo server and upload our prepared dataset to the 'ProPass' project.

```{r opal-setup, cache=FALSE}
library(opalr)
library(tibble)

upload_testing_dataset_table <- function(opal, project_name, table_name, local_file_path, file_type='csv') {
  if (! opal.project_exists(opal, project_name))
    opal.project_create(opal, project_name, database = "mongodb")
  
  if(file_type=='rda'){
    dataset_name <- load(file = local_file_path)
    dataset <- eval(as.symbol(dataset_name))
  }
  if(file_type=='csv'){
    dataset <- read.csv(file = local_file_path)
  }
  
  data <- as_tibble(dataset, rownames = '_row_id_')
  data$`_row_id_` <- as.numeric(data$`_row_id_`)
  
  opal.table_save(opal = opal, data, project_name, table_name, id.name = "_row_id_", force = TRUE)
}

opal <- opal.login('administrator','password', url='https://opal-demo.obiba.org', 
                   opts = list(ssl_verifyhost=0, ssl_verifypeer=0))
upload_testing_dataset_table(opal, project_name = 'ProPass', table_name = 'life1', file_type='rda', local_file_path='df.rda')
```

#### Server Configuration

We configure the Opal server for DataSHIELD analysis by setting general configuration options, installing required packages, and configuring DataSHIELD profiles. This includes installing the survival analysis packages from GitHub, setting up the DataSHIELD profile with all necessary packages, and configuring privacy and seed settings for reproducible analysis.

```{r server-config, results='hide', cache=FALSE}
config_json <- '{
  "name": "Opal",
  "defaultCharSet": "ISO-8859-1", 
  "enforced2FA": false,
  "allowRPackageManagement": true
}'

opal.put(
  opal, "system", "conf", "general",
  body = config_json,
  contentType = "application/json"
)

opalr::dsadmin.install_github_package(opal, "dsSurvival", "datashield", "v2.3.0-phase1") # This operation takes some minutes to complete
opalr::dsadmin.install_github_package(opal, "dsCoda", "datashield", "main")
opalr::dsadmin.install_package(opal, "dsTidyverse")
opalr::dsadmin.profile_init(opal, "default", c("dsBase", "dsSurvival", "dsTidyverse", "resourcer", "dsCoda"))
opalr::dsadmin.set_option(opal, "datashield.privacyControlLevel", "permissive")
opalr::dsadmin.set_option(opal, "	datashield.seed", "239")
opal.logout(opal)
```

### DataSHIELD Analysis

#### Connection and Initial Data Processing

We establish the DataSHIELD connection to our Opal server using the login builder pattern. This replicates the approach from the original `Example_script.html` script for ProPASS. The connection assigns our uploaded dataset to the 'D' symbol for analysis.

```{r datashield-connection, cache=FALSE}
builder <- DSI::newDSLoginBuilder()
builder$append(server = "study1", url = "https://opal-demo.obiba.org", user = "administrator", password = "password", table = "ProPass.life1", driver = "OpalDriver", options='list(ssl_verifyhost=0, ssl_verifypeer=0)')
logindata <- builder$build()

connections <- datashield.login(logins = logindata, assign = TRUE, symbol = "D")
```

#### Data wranggling

We will create a new variable `lipa` by adding the `lpa` and `stand` variables. Then we will select the variables `vpa`, `mpa`, `lipa`, `sleep`, and `sb` from the data.frame.

```{r data-wrangling, cache=FALSE}
ds.dim("D")
ds.mutate(
  df.name = "D",
  tidy_expr = list(lipa = lpa + stand),
  newobj = "D2"
)

ds.select(df.name = "D2", tidy_expr = list(vpa, mpa, lipa, sleep, sb), newobj = "D3")
```



#### lrEM

We will pass the subseted data frame to the `lrEM` function. We will use the `label` parameter to denote the missing values. We will use the `dl` parameter to denote the detection limits.

```{r lrEM, results='hide', cache=FALSE}
ds.lrEM(
  X = "D3",
  label = 0,
  dl = c(0.1667, 0.1667, 0.1667, 0.1667, 0.1667),
  objectname = "D4"
)
```

### Summary

This task 17 analysis demonstrates the complete DataSHIELD workflow to use the `lrEM` function, we included:

1. **Data Preparation**: Local data processing, server setup, and remote data preparation
2. **Data wrangling**: Create a new variable `lipa` by adding the `lpa` and `stand` variables. Then we will select the variables `vpa`, `mpa`, `lipa`, `sleep`, and `sb` from the data.frame.
3. **lrEM**: Impute the missing values in the data.frame.

The analysis successfully implements privacy-preserving statistical methods through DataSHIELD while maintaining analytical rigor comparable to traditional centralized approaches.

## Results

The `ds.lrEM` returns a new data.frame.

```{r results, cache=FALSE}
ds.class("D4")
```
