---
title: "Prerequisites"
format: html
engine: knitr
---

```{css}
#| echo: false
p {
  text-align: justify
}
```

This page outlines the software and knowledge requirements for the ProPASS DataSHIELD training materials.

## Software Requirements

### R and RStudio

You will need:

- **R** version 4.3.0 or later
  - Download from [CRAN](https://cran.r-project.org/)
- **RStudio Desktop** (recommended)
  - Download from [Posit](https://posit.co/download/rstudio-desktop/)

### Required R Packages

Install the following R packages:

```{r}
#| eval: false

# Core DataSHIELD packages
install.packages("DSI")
install.packages("DSOpal")

# DataSHIELD client packages
install.packages("devtools")
devtools::install_github("datashield/dsBaseClient", force = TRUE)

# Tidyverse-style data manipulation (used in ProPASS scripts)
devtools::install_github("molgenis/dsTidyverseClient")

# Survival analysis (for Module 2)
devtools::install_github("datashield/dsSurvivalClient")

# Helper packages
devtools::install_github("timcadman/ds-helper")

# Meta-analysis (for pooling results)
install.packages("metafor")
```

### Verify Installation

Run the following to confirm your packages are installed correctly:

```{r}
#| eval: false

library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsTidyverseClient)

# Check versions
packageVersion("dsBaseClient")
packageVersion("dsTidyverseClient")
```

## Server Access

For this training, we will use the public **Opal demo server** provided by OBiBa:

| Setting | Value |
|---------|-------|
| **Server URL** | `https://opal-demo.obiba.org` |
| **Username** | `dsuser` |
| **Password** | `P@ssw0rd` |
| **Profile** | `margin-idiom` |

::: {.callout-note}
## Public Demo Server
These credentials are publicly available for demonstration and training purposes.
:::

### Testing Your Connection

Test your connection with the demo server:

```{r}
#| eval: false

library(DSI)
library(DSOpal)

# Build login credentials for the demo server
builder <- DSI::newDSLoginBuilder()
builder$append(
  server = "demo",
  url = "https://opal-demo.obiba.org",
  user = "dsuser",
  password = "P@ssw0rd",
  table = "CNSIM.CNSIM1",
  profile = "margin-idiom"
)

logindata <- builder$build()

# Test connection
conns <- datashield.login(logins = logindata, assign = TRUE, symbol = "D")

# Verify connection
ds.ls()
ds.dim("D")

# Clean up
datashield.logout(conns)
```

## Knowledge Prerequisites

This workshop assumes familiarity with:

### R Programming
- Data manipulation with base R or tidyverse
- Writing and calling functions
- Working with data frames

### Statistics
- Descriptive statistics (mean, median, standard deviation)
- Linear and logistic regression
- Basic survival analysis concepts

### DataSHIELD Basics (Helpful but not required)
- Understanding of federated analysis concepts
- Basic DataSHIELD workflow (login, assign, analyze, logout)

## Additional Resources

- [DataSHIELD Wiki](https://data2knowledge.atlassian.net/wiki/spaces/DSDEV/overview)
- [dsBaseClient Reference Manual](https://cran.datashield.org/web/packages/dsBaseClient/dsBaseClient.pdf)
- [DataSHIELD Training Materials](https://github.com/isglobal-brge/workshop_datashield)

