---
title: "Filtering & Subsetting"
format: html
engine: knitr
---

```{css}
#| echo: false
p {
  text-align: justify
}
```

This chapter covers techniques for filtering and subsetting datasets in DataSHIELD based on specified conditions.

## Overview

In DataSHIELD, you cannot directly view individual-level data, but you can create subsets based on conditions. We use the tidyverse-style functions from `dsTidyverseClient`:

- `ds.filter()` - Filter rows based on conditions
- `ds.select()` - Select specific columns
- `ds.arrange()` - Sort data by columns
- `ds.dataFrameSubset()` - Subset with comparison operators (from dsBaseClient)

```{r}
#| label: setup
#| message: false
#| warning: false

# Load required packages
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsTidyverseClient)

# Connect to the server
builder <- DSI::newDSLoginBuilder()
builder$append(
  server = "demo",
  url = "https://opal-demo.obiba.org",
  user = "dsuser",
  password = "P@ssw0rd",
  table = "CNSIM.CNSIM1",
  profile = "lemon-donkey"
)
logindata <- builder$build()
conns <- datashield.login(logins = logindata, assign = TRUE, symbol = "D")
```

## Filtering Rows with ds.filter()

Filter rows based on conditions using familiar tidyverse syntax:

```{r}
# Filter for participants with BMI >= 25
ds.filter(
  df.name = "D",
  tidy_expr = list(D$PM_BMI_CONTINUOUS >= 25),
  newobj = "D_overweight"
)

# Check how many rows remain
ds.dim("D_overweight")
```

### Multiple Conditions

```{r}
# Filter for BMI between 20 and 35
ds.filter(
  df.name = "D",
  tidy_expr = list(D$PM_BMI_CONTINUOUS >= 20, D$PM_BMI_CONTINUOUS <= 35),
  newobj = "D_bmi_range"
)

ds.dim("D_bmi_range")
```

## Selecting Columns with ds.select()

Select specific columns from your data frame:

```{r}
# Select variables needed for analysis
ds.select(
  df.name = "D",
  tidy_expr = list(PM_BMI_CONTINUOUS, LAB_TSC, LAB_HDL, LAB_TRIG, GENDER, DIS_DIAB),
  newobj = "D_selected"
)

# Check the result
ds.colnames("D_selected")
```

## Sorting Data with ds.arrange()

Sort your data by one or more columns:

```{r}
# Sort by BMI
ds.arrange(
  df.name = "D",
  tidy_expr = list(PM_BMI_CONTINUOUS),
  newobj = "D_sorted"
)

ds.dim("D_sorted")
```

## Subsetting with ds.dataFrameSubset()

For subsetting based on comparison operators, use `ds.dataFrameSubset()`:

```{r}
# Keep only participants with BMI >= 25
ds.dataFrameSubset(
  df.name = "D",
  V1.name = "D$PM_BMI_CONTINUOUS",
  V2.name = "25",
  Boolean.operator = ">=",
  newobj = "D_bmi25plus"
)

# Check how many participants remain
ds.dim("D_bmi25plus")
```

### Available Operators

| Operator | Description |
|----------|-------------|
| `==` | Equal to |
| `!=` | Not equal to |
| `<` | Less than |
| `<=` | Less than or equal |
| `>` | Greater than |
| `>=` | Greater than or equal |

## Complete Case Analysis

Remove rows with any missing values:

```{r}
ds.completeCases(
  x1 = "D",
  newobj = "D_complete"
)

ds.dim("D_complete")
```

## ProPASS Script Workflow

::: {.callout-note}
## ProPASS Data
The ProPASS analysis uses different variables than the CNSIM demo data. Here's the workflow you'll use with ProPASS data:
:::

```{r}
#| eval: false

# 1. Sort data by ID
ds.arrange(df.name = "D", tidy_expr = list(ID), newobj = "D")

# 2. Filter out missing alcohol values
ds.filter(df.name = "D", tidy_expr = list(D$alc != 'NA'), newobj = "D")

# 3. Select variables needed for analysis
ds.select(
  df.name = "D",
  tidy_expr = list(ID, age, sex, smoke, edu, bmi, alc, med_combined, diet),
  newobj = "D2"
)

# 4. Filter for adequate follow-up (survival analysis)
ds.dataFrameSubset(
  df.name = "D",
  V1.name = "D$fup",
  V2.name = "1",
  Boolean.operator = ">=",
  newobj = "dat_ACM_BMI"
)
```

## Best Practices

::: {.callout-tip}
## Subsetting Guidelines
1. **Sort your data first**: Use `ds.arrange()` to ensure consistent ordering
2. **Filter early**: Remove missing/invalid data before other operations
3. **Check dimensions**: Always verify sample sizes after filtering
4. **Document your filters**: Keep track of exclusion criteria
5. **Consider disclosure**: Very small subsets may fail privacy checks
:::

## Exercise

Using the CNSIM data:

1. Filter for participants with BMI between 20 and 40 using `ds.filter()`
2. Select only the lab variables (LAB_TSC, LAB_HDL, LAB_TRIG) using `ds.select()`
3. Create a subset with cholesterol < 7 using `ds.dataFrameSubset()`

Check the sample size at each step.

```{r}
#| include: false

# Cleanup
datashield.logout(conns)
```

## Next Steps

Continue to [Conditional Operations](4-conditional-operations.qmd) to learn about creating conditional variables.
