---
title: "Compositional Data Analysis"
format: html
engine: knitr
---

```{css}
#| echo: false
p {
  text-align: justify
}
```

This chapter covers Compositional Data Analysis (CoDA) in DataSHIELD using the `dsCoda` package, which is essential for analyzing time-use data, nutrient compositions, and other proportional data.

## Overview

### What is Compositional Data?

Compositional data consists of parts that sum to a constant (e.g., 100% or 24 hours). Examples include:

- **Time-use data**: Sleep, sedentary time, light activity, moderate-to-vigorous activity
- **Nutrient composition**: Protein, fat, carbohydrates
- **Microbiome data**: Relative abundances of bacterial taxa
- **Body composition**: Fat mass, lean mass, bone mass

### Why Special Methods?

Standard statistical methods assume independence between variables. However, compositional parts are inherently dependent—if one increases, others must decrease. CoDA uses log-ratio transformations to handle this constraint properly.

## CoDA Concepts

### The Simplex

Compositional data lives in a constrained space called the simplex:

$$\mathcal{S}^D = \left\{ \mathbf{x} = (x_1, \ldots, x_D) : x_i > 0, \sum_{i=1}^D x_i = \kappa \right\}$$

where $\kappa$ is the constant sum (e.g., 24 hours for daily time-use).

### Log-Ratio Transformations

Three main transformations move compositional data from the simplex to unconstrained Euclidean space:

1. **ALR** (Additive Log-Ratio): Simple but asymmetric
2. **CLR** (Centered Log-Ratio): Symmetric but singular
3. **ILR** (Isometric Log-Ratio): Optimal for regression - **used in ProPASS**

### The ILR Transformation

The ILR transformation maps D-part compositional data to (D-1) dimensional Euclidean space:

$$\mathbf{z} = \log(\mathbf{x}) \cdot \mathbf{V}$$

where $\mathbf{V}$ is an orthonormal basis matrix constructed from a Sequential Binary Partition (SBP).

## Setup

::: {.callout-important}
## Package Requirements
CoDA analysis requires the `dsCoda` server-side package. Install the client package from GitHub.
:::

### Installing dsCodaClient

```{r}
#| eval: false

# Install from GitHub
devtools::install_github("datashield/dsCodaClient")
```

### Loading Libraries and Connecting

```{r}
#| label: setup
#| message: false
#| warning: false

# Load required packages
library(DSI)
library(DSOpal)
library(dsBaseClient)
library(dsTidyverseClient)
library(dsCodaClient)

# Connect to ProPASS data
builder <- DSI::newDSLoginBuilder()
builder$append(
  server = "study1",
  url = "https://opal-demo.obiba.org",
  user = "administrator",
  password = "password",
  table = "ProPass.df",
  profile = "margin-idiom"
)

logindata <- builder$build()
conns <- datashield.login(logins = logindata, assign = TRUE, symbol = "D")
```

### Exploring the Data

```{r}
# Check data structure
ds.colnames("D")
ds.dim("D")
```

## The ProPASS CoDA Workflow

The complete workflow for compositional data analysis follows these steps:

1. **Data Wrangling**: Combine activity categories and select compositional parts
2. **Handle Zeros**: Use `ds.lrEM()` to impute zeros below detection limits
3. **Create Composition**: Use `ds.acomp()` to create compositional objects
4. **Build Basis Matrix**: Use `build_sequential_ilr_V()` for ILR transformation
5. **Apply ILR**: Use `ds.ilr()` to transform to Euclidean space
6. **Combine with Covariates**: Merge ILR coordinates with outcome and covariate data
7. **Survival Analysis**: Fit Fine-Gray competing risks model with ILR coordinates

### Step 1: Data Wrangling

First, prepare the compositional parts. In ProPASS, light intensity physical activity (`lipa`) is created by combining `lpa` and `stand`:

```{r}
# Create combined light activity variable (lipa = lpa + stand)
ds.mutate(
  df.name = "D",
  tidy_expr = list(lipa = lpa + stand),
  newobj = "D2"
)

# Select the 5 compositional parts for analysis
# Order matters for ILR interpretation!
ds.select(
  df.name = "D2",
  tidy_expr = list(vpa, mpa, lipa, sleep, sb),
  newobj = "D3"
)

# Check the composition
ds.colnames("D3")
ds.dim("D3")
```

::: {.callout-note}
## Composition Parts
The 5-part composition represents daily time-use:
- **vpa**: Vigorous physical activity
- **mpa**: Moderate physical activity  
- **lipa**: Light intensity physical activity (lpa + stand)
- **sleep**: Sleep time
- **sb**: Sedentary behavior
:::

### Step 2: Handle Zeros with ds.lrEM()

Compositional data requires all parts > 0. The `ds.lrEM()` function uses log-ratio Expectation-Maximization to impute zeros and values below detection limits:

```{r}
#| results: hide

# Impute zeros using log-ratio EM algorithm
ds.lrEM(
  X = "D3",
  label = 0,                                         # Value indicating zeros/missing
  dl = c(0.1667, 0.1667, 0.1667, 0.1667, 0.1667),    # Detection limits (10 min = 0.1667 hours)
  objectname = "D4"
)
```

::: {.callout-tip}
## Understanding ds.lrEM() Parameters

- **X**: Data frame containing compositional parts
- **label**: Value that indicates zeros/missing (typically 0)
- **dl**: Detection limits for each part (minimum detectable value)
- **objectname**: Name for the output object

The detection limit of 0.1667 hours = 10 minutes represents the minimum meaningful activity bout.
:::

```{r}
# Verify the imputation
ds.class("D4")
ds.dim("D4")
```

### Step 3: Create Compositional Object with ds.acomp()

The `ds.acomp()` function creates an `acomp` class object that properly represents compositional data:

```{r}
#| results: hide

# Create acomp object
ds.acomp(
  X = "D4",
  objectname = "D5"
)
```

```{r}
# Verify the acomp object
ds.class("D5")
```

::: {.callout-note}
## What is an acomp object?
The `acomp` class (from the `compositions` R package) represents closed compositions where parts sum to a total. It enables proper compositional operations and transformations.
:::

### Step 4: Build the ILR Basis Matrix

The ILR transformation requires a basis matrix V. The `build_sequential_ilr_V()` function creates a sequential binary partition:

```{r}
# Get the number of compositional parts
n_parts <- ds.dim("D5")[[1]][2]
cat("Number of compositional parts:", n_parts, "\n")

# Build the sequential ILR basis matrix
V <- build_sequential_ilr_V(n_parts)

# View the basis matrix
print(V)
```

::: {.callout-note}
## Understanding the Basis Matrix
The basis matrix V defines how parts are contrasted in the ILR transformation:

- **Rows**: Original compositional parts (D parts)
- **Columns**: ILR coordinates (D-1 coordinates)
- **Values**: Weights for the log-ratio contrasts

For 5 parts, you get 4 ILR coordinates.
:::

### Step 5: Apply ILR Transformation with ds.ilr()

Apply the ILR transformation to convert the composition to unconstrained Euclidean coordinates:

```{r}
#| results: hide

# Apply ILR transformation
ds.ilr(
  X = "D5",
  V = V,
  objectname = "ilr_coords"
)
```

```{r}
# Check the result
ds.class("ilr_coords")
ds.dim("ilr_coords")
```

The ILR transformation creates D-1 = 4 coordinates from the 5-part composition.

### Step 6: Combine with Covariates for Analysis

To perform survival analysis, combine the ILR coordinates with outcome and covariate data:

```{r}
#| message: false
#| warning: false
#| output: false

# Load survival analysis package
library(dsSurvivalClient)

# Merge ILR coordinates back with original data
ds.dataFrame(
  x = c("D2", "ilr_coords"),
  newobj = "analysis_data"
)

# Convert covariates to appropriate types
ds.asFactor(input.var.name = "analysis_data$sex", newobj.name = "sex_factor")
ds.asNumeric(x.name = "analysis_data$age", newobj = "age_num")
ds.asFactor(input.var.name = "analysis_data$condition_competing_risk", newobj.name = "event_factor")
ds.asNumeric(x.name = "analysis_data$condition_fup", newobj = "fup_num")

# Combine into final analysis dataset
ds.dataFrame(
  x = c("analysis_data", "sex_factor", "age_num", "event_factor", "fup_num"),
  newobj = "dat_coda"
)

# Rename the ILR coordinate columns to meaningful names
# Based on sequential binary partition with order: vpa, mpa, lipa, sleep, sb
ds.rename(
  df.name = "dat_coda",
  tidy_expr = list(
    VPAvsAll = `ilr_coords`,        # VPA vs (mpa, lipa, sleep, sb)
    MPAvsRemaining = `NA.`,         # MPA vs (lipa, sleep, sb)
    LIPAvsRemaining = `NA..1`,      # LIPA vs (sleep, sb)
    SLEEPvsSB = `NA..2`             # Sleep vs SB
  ),
  newobj = "dat_coda"
)

# Select only the columns needed for survival analysis
# This avoids issues with variables that have too many missing values
ds.select(
  df.name = "dat_coda",
  tidy_expr = list(VPAvsAll, MPAvsRemaining, LIPAvsRemaining, SLEEPvsSB, age_num, sex_factor, event_factor, fup_num),
  newobj = "dat_coda_clean"
)
```

```{r}
# Check the analysis data
ds.colnames("dat_coda_clean")
ds.dim("dat_coda_clean")
```

### Step 7: Survival Analysis with ILR Coordinates

Now we fit a Fine-Gray competing risks model using the ILR coordinates as exposures:

```{r}
#| message: false
#| warning: false
#| output: false

# Setup datadist for rms functions
# Center ILR coordinates at the compositional mean (geometric center)
ds.datadist(
  data = "dat_coda_clean",
  adjust_to = list(
    age_num = "mean"
  )
)
ds.useDatadist(datadist = "datadist_dat_coda_clean")

# Create Fine-Gray object for competing risks analysis
# etype = "1" specifies the event of interest
ds.finegray(
  formula = "Surv(fup_num, event_factor) ~ .",
  data = "dat_coda_clean",
  etype = "1",
  newobj = "fg_coda"
)
```

```{r}
#| message: false
#| warning: false
#| output: false

# Fit Cox model on Fine-Gray data with ILR coordinates
# Each ILR coordinate represents a specific time-use contrast
ds.coxphSLMAassign(
  formula = "survival::Surv(fgstart, fgstop, fgstatus) ~ VPAvsAll + MPAvsRemaining + LIPAvsRemaining + SLEEPvsSB + age_num + sex_factor",
  dataName = "fg_coda",
  objectname = "cph_coda",
  use.rms = TRUE
)
```

```{r}
# View model results
ds.coxphSummary("cph_coda")
```

::: {.callout-note}
## ILR Coordinate Interpretation
The ILR coordinates in the Cox model represent log-ratio contrasts based on the sequential binary partition:

| Coordinate | Contrast | Interpretation |
|------------|----------|----------------|
| **VPAvsAll** | VPA vs (mpa, lipa, sleep, sb) | Effect of increasing VPA relative to all other activities |
| **MPAvsRemaining** | MPA vs (lipa, sleep, sb) | Effect of increasing MPA relative to lighter activities |
| **LIPAvsRemaining** | LIPA vs (sleep, sb) | Effect of increasing LIPA relative to inactive time |
| **SLEEPvsSB** | Sleep vs SB | Effect of increasing sleep relative to sedentary behavior |

A positive coefficient means higher hazard ratio when that ILR coordinate increases.
:::

## Basis Rotation

The order of compositional parts determines the ILR interpretation. **Rotation is simply done by reordering the variables in `ds.select()`** - this changes which component is contrasted first:

```{r}
#| eval: false

# ============================================================
# ROTATION IS CONTROLLED BY VARIABLE ORDER IN ds.select()
# ============================================================

# Rotation 1: VPA first (default in ProPASS)
ds.select(df.name = "D2", tidy_expr = list(vpa, mpa, lipa, sleep, sb), newobj = "comp_rot1")

# Rotation 2: Sleep first (to study sleep effects)
ds.select(df.name = "D2", tidy_expr = list(sleep, sb, lipa, mpa, vpa), newobj = "comp_rot2")

# Rotation 3: Sedentary behavior first
ds.select(df.name = "D2", tidy_expr = list(sb, sleep, lipa, mpa, vpa), newobj = "comp_rot3")
```

After selecting with a different order, apply the same workflow:

```{r}
#| eval: false

# Example: Sleep-first rotation
ds.select(df.name = "D2", tidy_expr = list(sleep, sb, lipa, mpa, vpa), newobj = "comp_rotated")

# Same workflow applies - order of dl must match new variable order!
ds.lrEM(
  X = "comp_rotated",
  label = 0,
  dl = c(0.1667, 0.1667, 0.1667, 0.1667, 0.1667),
  objectname = "comp_rotated_imp"
)

ds.acomp(X = "comp_rotated_imp", objectname = "comp_rotated_acomp")

# Build basis and transform (same function, interpretation changes)
V_rot <- build_sequential_ilr_V(5)
ds.ilr(X = "comp_rotated_acomp", V = V_rot, objectname = "ilr_rotated")
```

::: {.callout-tip}
## How Rotation Works
Rotation is **not** a separate function - it's simply the order you specify in `ds.select()`:

| Variable Order | First ILR Interprets |
|----------------|---------------------|
| `vpa, mpa, lipa, sleep, sb` | VPA vs. all other activities |
| `sleep, sb, lipa, mpa, vpa` | Sleep vs. all waking activities |
| `sb, sleep, lipa, mpa, vpa` | Sedentary vs. all other behaviors |

**Strategy**: Put your exposure of interest first to get its specific contrast in `ilr1`.
:::

## Interpreting CoDA Results

### ILR Coefficient Interpretation

ILR coefficients are not directly interpretable in terms of original parts. They represent:

- **ilr1**: Log-ratio contrast between first part and geometric mean of remaining parts
- **ilr2**: Log-ratio contrast within the remaining parts
- And so on...

### Isotemporal Substitution

To understand the effect of reallocating time between activities, use isotemporal substitution analysis. This is typically computed client-side using the ILR coefficients from the model.

## Summary of Key Functions

| Function | Purpose | Input | Output |
|----------|---------|-------|--------|
| `ds.lrEM()` | Impute zeros | Data frame with composition | Data frame with imputed values |
| `ds.acomp()` | Create composition object | Data frame | acomp object |
| `build_sequential_ilr_V()` | Build basis matrix | Number of parts (integer) | Matrix V |
| `ds.ilr()` | ILR transformation | acomp object + V matrix | ILR coordinates |

## Best Practices

::: {.callout-tip}
## CoDA Analysis Tips

1. **Handle zeros properly**: Always use `ds.lrEM()` before transformation - zeros break log-ratios
2. **Check closure**: Ensure compositional parts sum to expected total (e.g., 24 hours)
3. **Choose rotation wisely**: Put your exposure of interest first for clearer interpretation
4. **Set appropriate detection limits**: Use meaningful minimum values (e.g., 10 minutes for activities)
5. **Report geometric means**: Present compositional means correctly (not arithmetic means)
:::

```{r}
#| include: false

# Cleanup
datashield.logout(conns)
```

## References

- Dumuid, D. et al. (2018). Compositional data analysis for physical activity, sedentary time and sleep research. *Statistical Methods in Medical Research*.
- Chastin, S. et al. (2015). Combined effects of time spent in physical activity, sedentary behaviors and sleep on obesity and cardio-metabolic health markers.
- [compositions R package](https://cran.r-project.org/package=compositions)
- [zCompositions R package](https://cran.r-project.org/package=zCompositions) (for lrEM algorithm)

## Next Steps

Continue to the next chapter or return to [Survival Analysis](8-survival-analysis.qmd) to combine CoDA with time-to-event outcomes.
